---
title: "An Analysis of NBA Player Salary Variation Across Different Player Statistics" 
author: "The Significant Others - Siri,Marin,ReNata"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \onehalfspacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE, comment=NA, warnings=FALSE,message=FALSE)
library(eeptools)
```

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## Use this chunk for reading in packages.
## Use this chunk for reading in the data, combining the data, and any other major data cleaning that is not relevant to show.
## Remove the ## before all code.
```

\newpage

## Introduction

1)  Do NBA players in the Western Conference earn higher average salaries in following season compared to players in the Eastern Conference?
2)  Do NBA players who average more minutes of playing time per game in a season earn higher salaries than players who average fewer minutes per game?
3)  Do NBA player salaries vary across age groups (young, prime, veteran, and old)?

## Data Summary

### Metadata Description

### EDA Summary

```{r}
nba_data <- read.csv("https://raw.githubusercontent.com/sirinarne/Stat3220_Final/refs/heads/main/NBA_Data_Final1.csv", header = T)

#Variable Recoding
#Age --> Split into groups
nba_data$AGE_GROUP <- cut(nba_data$AGE, breaks = c(18,24,29,35,Inf), labels = c("Young", "Prime", "Veteran", "Old"), right=FALSE)

#Team --> Conference
east_teams <- c("ATL","BOS","BKN","CHA","CHI","CLE","DET","IND","MIA","MIL","NYK","ORL","PHI","TOR","WAS")
nba_data$CONFERENCE <- ifelse(nba_data$TEAM %in% east_teams, "East", "West")


#Net Rating --> Tiers
nba_data$NETRTG_TIER <- cut(nba_data$NETRTG,breaks = quantile(nba_data$NETRTG, probs = c(0, 0.33, 0.66, 1)),labels = c("Low Impact", "Average", "High Impact"),include.lowest = TRUE)


nba_data$SALARY<-decomma(nba_data$SALARY)
nba_data$SALARY<-nba_data$SALARY/1000000
```

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## Use this code chunk can be used to execute any functions or summaries 
## that will not be included in your output such as correlations or 5 number summaries

#five number summaries 
tapply(nba_data$SALARY,nba_data$AGE_GROUP,summary)
tapply(nba_data$SALARY,nba_data$CONFERENCE,summary)
tapply(nba_data$SALARY,nba_data$NETRTG_TIER,summary)


#correlations 
#Correlations
corPIE<-cor(nba_data$PIE,nba_data$SALARY) 
corPCTTS<-cor(nba_data$TS_PCT,nba_data$SALARY) 
corMIN<-cor(nba_data$MIN,nba_data$SALARY) 
corOFFERTG<-cor(nba_data$OFFRTG,nba_data$SALARY) 
corPCTUSG<-cor(nba_data$USG_PCT,nba_data$SALARY)  

corPIE
corPCTTS
corMIN
corOFFERTG
corPCTUSG
```

Response Variable (Salary): The distribution of NBA player salaries for the season 2024-2025 ranges between 20 and 60 million dollars. The distribution indicates an overall downward trend(skewed right), with most players earning between 20 and 30 million. The distribution is also unimodal and there are no extreme outliers.

Quantitative EDA Summary:

The Player Impact Estimate seems to have the strongest correlation with a player’s salary the at 0.59. True shooting percentage, minutes played/game,and usage percentage have lower correlations with a player’s salary of 0.49, 0.45, and 0.47 respectively. The weakest correlation is between offensive rating and player salary at 0.36.

Qualitative EDA Summary:

There appears to be a relationship between age group and player salary because the mean value of y is different at each level. Players that belonged to the “Old” age group in the previous season earned the highest salary (44.67), while young and prime players have relatively similar average earnings (33.59 and 32.46), and veteran players earned an average of 36.57.

There appears to be a relationship with conference and player salary because the mean value of y is different for each level. Players in the Eastern conference had an salary of 32.70, while the Western conference players had a higher average salary of 36.49.

There appears to be a relationship between the net rating tier of a player and their salary because the mean value of y increases across levels. Low impact players have the lowest mean salary of 29.01, while average players have an average salary of 36.05, and high impact players have the highest average salary of 38.89.

Interactions Summary:

The interaction plot shows that there may be an interaction between net rating tier and age group on salary. The slopes do not appear to be parallel and there is some crossing over that occurs, indicating that the effect of net rating tier on salary may be dependent on age.

The grouped scatterplot indicates a generally positive correlation between player impact estimate and salary. The effect of a player's impact estimate on salary does not seem to be dependent on age as the slopes are generally similar.

There may be an interaction between the average playing time/per and net rating tier. The slopes are fairly different as the high impact slope is the steepest, meaning that the highest impact players who play for a longer period of time per game earn the highest salaries. Because the lines are not parallel, it serves as evidence of a potential interaction.

## Analysis and Results

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## This code chunk can be used to execute any functions or summaries that will not be included in your output

```

## Conclusions

\newpage

## Appendix A: Data Dictionary

| Variable Name | Abbreviated Name | Description |
|:----------------------:|:----------------------:|:----------------------:|
| Salary | SALARY | Player salary in millions of dollars |
| Player Impact Estimate | PIE | Player's overall contribution to the game; expressed as a percentage |
| True Shooting Percentage | TS_PCT | Measure player's efficiency at scoring (percentage) |
| Minutes Played | MIN | Average minutes played per game |
| Offensive Rating | OFFRTG | Estimates offensive efficiency by calculating points scored per 100 possessions |
| Usage Percentage | USG_PCT | Estimates the percentage of a team's possessions a player uses while on the court (percentage of possessions) |
| Net Rating Tier | NETRTG_TIER | Derived from quantitative variable (NETRTG); classifies players into three groups: low impact, average, or high impact |
| Conference | CONFERENCE | Indicates whether a player is in the Eastern or Western conference |
| Age Group | AGE_GROUP | Grouping of players into young, prime, veteran, or old based on age |

\newpage

## Appendix B: Data Rows

```{r, echo=FALSE}
# Use head() function to output first several rows
head(nba_data)
```

\newpage

### Appendix C: Exploratory Data Analysis

```{r, echo=FALSE}
## Use this code  to include the relevant plots from EDA
## This should NOT be a dump of all of your plots and outputs.
## Refer to the Rmarkdown guide for sizing information on your plots

#Histogram of Response
options(scipen=10)
hist(nba_data$SALARY, xlab="NBA Player Salary", main="Distribution of NBA Player Salaries") 

#Quantitative
plot(nba_data$PIE, nba_data$SALARY,xlab="Player Impact Estimate",ylab="Player Salary")
plot(nba_data$TS_PCT, nba_data$SALARY,xlab="True Shooting %",ylab="Player Salary ")
plot(nba_data$MIN, nba_data$SALARY,xlab="Minutes Played/Game",ylab="Player Salary ")
plot(nba_data$OFFRTG,nba_data$SALARY,xlab="Offensive Rating", ylab="Player Salary ")
plot(nba_data$USG_PCT, nba_data$SALARY,xlab="Usage Percentage", ylab="Player Salary")

#Qualitative
boxplot(SALARY~ NETRTG_TIER,nba_data, ylab="Player Salary")
boxplot(SALARY~CONFERENCE,nba_data, ylab="Player Salary")
boxplot(SALARY~ AGE_GROUP,nba_data, ylab="Player Salary")

#Interactions
# qual x qual 
interaction.plot(nba_data$NETRTG_TIER, nba_data$AGE_GROUP, nba_data$SALARY,fun=mean,trace.label="Age Group", xlab="Impact Level",ylab="Salary")


#quant x qual
plot(x=nba_data$PIE, y=nba_data$SALARY,pch=as.numeric(factor(nba_data$AGE_GROUP)),xlab="Player Impact Estimate",ylab="Salary") 
legend("topright",legend = levels(factor(nba_data$AGE_GROUP)),
       pch = 1:nlevels(factor(nba_data$AGE_GROUP)))
for (g in levels(as.factor(nba_data$AGE_GROUP))) {
  group_subset <- subset(nba_data, AGE_GROUP == g)
  model <- lm(SALARY ~ PIE, data = group_subset)
  abline(model, lwd = 2) 
}


# quant x qual
plot(x=nba_data$MIN, y=nba_data$SALARY,pch=as.numeric(factor(nba_data$NETRTG_TIER)),xlab="Minutes Played",ylab="Salary") 
legend("topright",legend = levels(factor(nba_data$NETRTG_TIER)),
       pch = 1:nlevels(factor(nba_data$NETRTG_TIER)))
for (g in levels(as.factor(nba_data$NETRTG_TIER))) {
  group_subset <- subset(nba_data, NETRTG_TIER == g)
  model <- lm(SALARY ~ MIN, data = group_subset)
  abline(model, lwd = 2)
}


# This partition is optional, but can be useful for a simple way to resize your plots
par(mfrow=c(2,3),pin=c(1,1),mar=c(7.1,4.1,3,1))
```

\newpage

## Appendix D: Final Model Output and Plots

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## This code chunk can be used to execute any functions or summaries that will not be included in your output
```

```{r, echo=FALSE}
## Use this section to output the summary of your final model
## Include the relevant plots from assumptions and your additional techniques
## This should NOT be a dump of all of your plots and outputs.
## Refer to the Rmarkdown guide for sizing information on your plots
```

\newpage

## Appendix E: References

### Background

### Data Sources

National Basketball Association. (n.d.). *Players — Advanced*. NBA.com. Retrieved April 11, 2026, from <https://www.nba.com/stats/players/advanced?sort=W&dir=-1&Season=2023-24>

ESPN. (2025). *NBA salaries — 2025*. Retrieved April11, 2026, from <https://www.espn.com/nba/salaries/_/year/2025>

### Additional Help
